---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}

#setwd("~/R/reproResearch/RepData_PeerAssessment1")
library(ggplot2)
library(dplyr)

unzip('activity.zip')
activity <- read.csv("activity.csv")
str(activity)
```
Coerce the 'date' column from character to Date type
```{r}
activity[,2] <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?
```{r}

byDay <- summarize(group_by(activity, date), totalSteps=sum(steps))
par(mfrow = c(1,1))
hist(byDay$totalSteps, main = "Total number of Steps in a Day", xlab = "Steps in a day")

```
We find the mean/median by day and create a table with the results, we see lots of NA or 0's.

```{r}
meanByDay <- summarize(group_by(activity, date), meanSteps = mean(steps)) 
 
meanByDay$meanSteps
```

```{r}
medianByDay <- summarize(group_by(activity, date), medianSteps = median(steps))
medianByDay$medianSteps
```

## What is the average daily activity pattern?
Create plots showing the average/median number of steps by time interval
```{r}
byInterval <-summarize(group_by(activity, interval ), averageSteps = mean(steps, na.rm = T), median = median(steps, na.rm = T))

par(mfrow = c(2,1))
plot(byInterval$interval, byInterval$averageSteps, type = 'l', ylab = "Mean number of Steps", xlab = "Time interval")
plot(byInterval$interval, byInterval$median, type = 'l', ylab = "Median number of Steps", xlab = "Time interval")
```

Let's find the interval with the max number of average steps
```{r}
byInterval[byInterval$averageSteps == max(byInterval$averageSteps),]

```


## Imputing missing values
I think replacing by average value by interval will give us a fair guestimate.  Average by day is unsatisfactory since some days are composed entirely of missing values, such as the first day.  Replacingby median by interval results in many 0 values, which 
isn't as interesting?

```{r}

newActivity <- activity
for (int in newActivity$interval){
    tempMean <- mean(newActivity$steps[newActivity$interval == int], na.rm=T)
    newActivity[newActivity$interval == int & is.na(newActivity)] <- tempMean
}
summary(newActivity)

```

Creating a histogram of the total number of steps by day of our NA free newActivity data
```{r}

newbyDay <- summarize(group_by(newActivity, date), totalSteps=sum(steps))

hist(newbyDay$totalSteps,main = 'Histogram of Total steps by day', xlab = "Total step by day")
newActivity[3,]
```


Create plots showing the average/median number of steps by time interval
```{r}
newbyInterval <-summarize(group_by(newActivity, interval ), averageSteps = mean(steps, na.rm = T), median = median(steps, na.rm = T))

par(mfrow = c(2,1))
plot(newbyInterval$interval, newbyInterval$averageSteps, type = 'l', ylab = "Mean number of Steps", xlab = "Time interval")
plot(newbyInterval$interval, newbyInterval$median, type = 'l', ylab = "Median number of Steps", xlab = "Time interval")
```


A list of the mean number of steps by day
```{r}
newMeanByDay <- summarize(group_by(newActivity, date), meanSteps = mean(steps)) 
newMeanByDay$meanSteps

```

a list of the median number of steps by day
```{r}
newMedianByDay <- summarize(group_by(newActivity, date), medianSteps = median(steps))
newMedianByDay$medianSteps

```


Overall we see that imputing the inverval mean for the missing data has increased the average number of steps on only a few days.
The medians again stayed mostly unchanged.  The NA medians have changed to 34.11.  If the median was 0 before, then it continues to be 0.



## Are there differences in activity patterns between weekdays and weekends?
Create a new "day" column containing factor variables inticating whether the day was a weekday or weekend.

```{r}
newActivity <- mutate(newActivity, day = weekdays(date))
newActivity[newActivity$day == "Saturday"| newActivity$day == 'Sunday',4] <- 'weekend'
newActivity[newActivity$day != 'weekend', 4] <- "weekday"
newActivity$day <- factor(newActivity$day, levels = c("weekend", "weekday"))
```

Plot the average number of steps by interval for both weekdays and weekends.

```{r}
bydayInterval <- summarize(group_by(newActivity,day, interval), steps = mean(steps))
qplot(interval, steps, data =bydayInterval, facets =day~., geom = 'line' )