---
title: "Project1"
author: "Dave Lofte"
date: "Monday, February 09, 2015"
output: html_document
---



###Loading in the data

```{r}


library(ggplot2)
library(dplyr)

unzip('activity.zip')
activity <- read.csv("activity.csv")
str(activity)
```

Coerce the 'date' column from character to Date type
I also create a time column containing both the date and interval
```{r}
activity[,2] <- as.Date(activity$date)
summary(activity[,2])
time <- paste(activity$date, activity$interval)
```

What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.
1. Make a histogram of the total number of steps taken each day
2. Calculate and report the mean and median total number of steps taken per day
```{r}
group_by(activity,date, interval)
byDay <- summarize(group_by(activity, date), totalSteps=sum(steps))
par(mfrow = c(1,1))
hist(byDay$totalSteps, main = "Total number of Steps in a Day", xlab = "Steps in a day")

```

We find the mean/median by day and create a table with the results, we see lots of NA or 0's.

```{r}
meanByDay <- summarize(group_by(activity, date), meanSteps = mean(steps)) 
medianByDay <- summarize(group_by(activity, date), medianSteps = median(steps)) 

results <- left_join(byDay, meanByDay)
results <- left_join(results, medianByDay)
results
```

Create plots showing the average/median number of steps by time interval
```{r}
byInterval <-summarize(group_by(activity, interval ), averageSteps = mean(steps, na.rm = T), median = median(steps, na.rm = T))

par(mfrow = c(2,1))
plot(byInterval$interval, byInterval$averageSteps, type = 'l')
plot(byInterval$interval, byInterval$median, type = 'l')
```

Let's find the interval with the max number of average steps
```{r}
byInterval[byInterval$averageSteps == max(byInterval$averageSteps),]

```


##Imputing Missing values
1. Calculate and report the total number of missing values in the data set
```{r}
summary(activity)
```
We see from the summary that all the NA values are located in the "steps" column
and that there are 2304 missing values or NA values

2.  I think replacing by average value
by interval will give us a fair guestimate.  Average by day is unsatisfactory since
some days are composed entirely of missing values, such as the first day.  Replacing
by median by interval results in many 0 values, which isn't as interesting?

3. I am creating a new data frame 'newActivity' with all the NA values replaced by the mean by interval.
```{r}

newActivity <- activity
for (int in newActivity$interval){
    tempMean <- mean(newActivity$steps[newActivity$interval == int], na.rm=T)
    newActivity[newActivity$interval == int & is.na(newActivity)] <- tempMean
}
summary(newActivity)
head(newActivity)

```

Creating a histogram of the total number of steps by day of our NA free newActivity data
```{r}
group_by(newActivity,date, interval)
newbyDay <- summarize(group_by(newActivity, date), totalSteps=sum(steps))

hist(newbyDay$totalSteps)
newActivity[3,]
```


```{r}
newMeanByDay <- summarize(group_by(newActivity, date), meanSteps = mean(steps)) 
newMedianByDay <- summarize(group_by(newActivity, date), medianSteps = median(steps)) 
summary(newMedianByDay)
str(newMeanByDay)


newResults <- left_join(newbyDay, newMeanByDay)
newResults <- left_join(newResults, newMedianByDay)
newResults
```

We compare the total of the Activity data.  Column 1 is the total # of steps from the data with imputed values for NAs.
Column 2 gives the total number of steps from the original data.
```{r}
cbind(sum(newResults$totalSteps), sum(results$totalSteps, na.rm = T))
```

Create a new "day" column containing factor variables inticating whether the day was a weekday or weekend.

```{r}
newActivity <- mutate(newActivity, day = weekdays(date))
newActivity[newActivity$day == "Saturday"| newActivity$day == 'Sunday',4] <- 'weekend'
newActivity[newActivity$day != 'weekend', 4] <- "weekday"
newActivity$day <- factor(newActivity$day, levels = c("weekend", "weekday"))
```

Plot the average number of steps by interval for both weekdays and weekends.

```{r}
bydayInterval <- summarize(group_by(newActivity,day, interval), steps = mean(steps))
qplot(interval, steps, data =bydayInterval, facets =day~., geom = 'line' )
?qplot
